Design of 1Mbit RRAM memory for NV applications

### 2014-05-21 Thesis/Paper/Paper_template.tex
paper geschreven

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
  %\graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation
% can be obtained at: 
% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found at:
% http://www.ctan.org/tex-archive/info/epslatex/
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.
%
% You can find documentation about the pdfTeX application at:
% http://www.tug.org/applications/pdftex

@@

@@

\section{Sense amplifier overlap techniques}
The sense amplifier used in this design is the drain-input latch-type SA (see figure \ref{ourSA}). Its input/output nodes are connected to the output nodes of the local block through complementary passgates. There are two ways to implement the latch timing cycle: one could seperate the pass operation from the latch operation or could allow overlap between these two operations.

\subsection{No overlap}
Using this control scheme, the SA would be separated from the local block when enabled. Offset voltage spread is mainly determined by the $\Delta_{V_{T}}$ and $\Delta_{beta}$ variations of the differential pairs of the SA. These contributions can be decreased by increasing the sizes of the differential pair transistors. Sizing up the top and bottom transistor has not more influence on the latching speed rather than the offset voltage. There is also a maybe slightly surprising contribution on the offset voltage by the passgates. This can be explained by the charge injection of the passgates: when the passgates are turned on, the output voltage of the local block is passed on through the input/output node of the SA almost perfectly - whether there are variations on the transistors or not. When the passgates are turned off, a charge injection occurs on the SA input/output node - distorting the original voltage.

% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.
% For figures, \caption should occur after the \includegraphics.
% Note that IEEEtran v1.7 and later has special internal code that
% is designed to preserve the operation of \label within \caption
% even when the captionsoff option is in effect. However, because
% of issues like this, it may be the safest practice to put all your
% \label just after \caption rather than within \caption{}.
%
% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
% option should be used if it is desired that the figures are to be
% displayed while in draft mode.
%
%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{myfigure}
% where an .eps filename suffix will be assumed under latex, 
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
%\caption{Simulation Results.}
%\label{fig_sim}
%\end{figure}

>>>>

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
  %\graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
% graphicx was written by David Carlisle and Sebastian Rahtz. It is
% required if you want graphics, photos, etc. graphicx.sty is already
% installed on most LaTeX systems. The latest version and documentation
% can be obtained at: 
% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
% Another good source of documentation is "Using Imported Graphics in
% LaTeX2e" by Keith Reckdahl which can be found at:
% http://www.ctan.org/tex-archive/info/epslatex/
%
% latex, and pdflatex in dvi mode, support graphics in encapsulated
% postscript (.eps) format. pdflatex in pdf mode supports graphics
% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
% which can result in "jaggedy"/blurry rendering of lines and letters as
% well as large increases in file sizes.
%
% You can find documentation about the pdfTeX application at:
% http://www.tug.org/applications/pdftexhttps://www.google.be/search?q=or+no&ie=utf-8&oe=utf-8&aq=t&rls=org.mozilla:en-US:official&client=firefox-a&gfe_rd=cr&ei=w3x8U6meBtCFOu37gLgF#q=variations+or+no&rls=org.mozilla:en-US:official

@@

\begin{figure}[ht!]
  \centering
  \includegraphics[width=0.5\textwidth]{../fig/paper-architecture.png}
  \caption{Overview architecture}
  \label{fig:architecture}
\end{figure}

@@

\section{Sense amplifier overlap techniques}
The sense amplifier used in this design is the drain-input latch-type SA (see figure \ref{fig:ourSA}). Its input/output nodes are connected to the output nodes of the local block through complementary passgates. There are two ways to implement the latch timing cycle: one could separate the pass operation from the latch operation or could allow overlap between these two operations.

\begin{figure}[ht!]
  \centering
  \includegraphics[width=0.25\textwidth]{../fig/hfdstk-sensamp-ourSA.png}
  \caption{Drain-input latch-type sense amplifier used in this design}
  \label{fig:ourSA}
\end{figure}

\subsection{No overlap}
Using this control scheme, the SA would be separated from the local block when enabled. Offset voltage spread is mainly determined by the $\Delta_{V_{T}}$ and $\Delta_{\beta}$ variations of the differential pairs of the SA. These contributions can be decreased by increasing the sizes of the differential pair transistors. Sizing up the top and bottom transistor has not more influence on the latching speed rather than the offset voltage. There is also a maybe slightly surprising contribution on the offset voltage by the passgates. This can be explained by the charge injection of the passgates: when the passgates are turned on, the output voltage of the local block is passed on to the input/output node of the SA almost perfectly - whether there are variations on the transistors or not. When the passgates are turned off, a charge injection occurs on the SA input/output node - distorting the original voltage. The SA operates differentially though, so as long as this charge injection is matched at the two input/output nodes there would be no problem. $\beta$ mismatch of the passgate transistors however results in charge injection mismatch (see figure \ref{fig:chargeinjection}). Hence the contribution of the passgates to the offset voltage spread. This mismatch can be reduced by sizing up the passgates transistors. More charge is injected when the passgates are turned off, but the difference at the two sides is reduced.

\begin{figure}[ht!]
  \centering
  \includegraphics[width=0.5\textwidth]{../fig/hfdstk-sensamp-chargeinjectionmismatch.png}
  \caption{Charge injection mismatch due to $\beta$ variations of passgate transistors}
  \label{fig:chargeinjection}
\end{figure}

\subsection{Overlap}
If the passgates are still turned on when the SA starts latching, the voltage difference of the input/output nodes has not experienced charge injection mismatch. When the voltage difference has been sufficiently amplified, the passgates are turned off. Charge injection (mismatch) will occur, but it will not change the outcome of the latching anymore. During the overlap of the pass and latching operation, the passgates can be modeled by resistors. Neglecting the resistance of the passgates in the local blocks, the situation can be depicted as in figure \ref{fig:RC-latch}. CL is the bitline capacitance.

\begin{figure}[ht!]
  \centering
  \includegraphics[width=0.25\textwidth]{../fig/hfdstk-sensamp-RC-latch.png}
  \caption{Simplified circuit when overlap between pass enable and latch enable is applied}
  \label{fig:RC-latch}
\end{figure}

One could suspect that a large BL capacitance would significantly increase the latching time. While this is true for small values of R, as can be seen in figure \ref{fig:RC-latch-sim}, for greater values the latching goes through two phases: during the first phase, it appears as if the big capacitance is decoupled from the SA. After this fast phase, settling is much slower.

\begin{figure}[ht!]
  \centering
  \includegraphics[width=0.5\textwidth]{../fig/hfdstk-sensamp-RC-latch-sim.png}
  \caption{Transient simulation for the schematic of figure \ref{fig:RC-latch} for different values of R, CL = 46fF, SA is minimal, no variations are included in simulation}
  \label{fig:RC-latch-sim}
\end{figure}

This RC-latch effect arises when the RC-product is large. When it occurs, the capacitance behaves as a short-circuit and current flows through the resistor and a corresponding voltage drop over the resistor builds up. Afterwards the load capacitance charges itself at a much lower time constant.

\paragraph{}
The overlap between the enabling of the passgates and the SA thus needn't be this great: a load-less SA latching delay suffices. For this situation $V_{T}$ and $\beta$ mismatch of the ;passgate transistors results in mismatch of the resistors. The interaction between these resistors and the SA is strongly non-linear. Depending on the precharged values of the load capacitances, this R mismatch can result in incorrect latching. There is thus still a contribution of the passgates to the offset voltage spread. The interaction between R and the SA reduces the contributions of the differential pair mismatch however. For a minimal SA, the offset voltage spread is smaller with overlap than without. By sizing all the transistors, this spread can be arbitrarily reduced. The resistance of the passgates mustn't be too small however, or slow settling might occur. Then again, a large SA would probably charge the load capacitance effortlessly anyway.
% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.
% For figures, \caption should occur after the \includegraphics.
% Note that IEEEtran v1.7 and later has special internal code that
% is designed to preserve the operation of \label within \caption
% even when the captionsoff option is in effect. However, because
% of issues like this, it may be the safest practice to put all your
% \label just after \caption rather than within \caption{}.
%
% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
% option should be used if it is desired that the figures are to be
% displayed while in draft mode.
%
%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{myfigure}
% where an .eps filename suffix will be assumed under latex, 
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
%\caption{Simulation Results.}
%\label{fig_sim}
%\end{figure}
